Generalized Universe Hierarchies and First-Class Universe Levels

### 2021-02-22 paper.tex
progress

Users of type theories often view universe levels as an unwieldy bureaucratic
detail, a necessary annoyance in service of boosting expressive power while
retaining logical consistency. However, universe hierarchies are not going away
any time soon in practical implementations of type theory. In recent
developments of systems, we are getting more universes and more adjacent
features:
\begin{itemize}
\item Agda recently added a limited cumulativity as an optional feature
  for universes \cite{agdadocs}, and the upcoming 2.6.2 version will extend the $\omega+1$
  universe hierarchy to $\omega*2$.
\item Coq added support for cumulative inductive types \cite{timany18cumulative}
  and a form of bounded universe polymorphism \cite{ziliani15unification}.
\end{itemize}
\noindent At this point, there is a veritable zoo of universe features in existing
implementations. We have perhaps even more design choices when considering the
formal metatheory of type theories. Do type formers stay in the same
universe, or take the $\sqcup$ of universes of constituent types? Can
eliminators target any universe, or do we instead use lifting operators to cross
levels? What kind of universe polymorphism do we have, can we quantify over
bounds? Is there a type of levels, or are levels in a separate syntactic layer?

@@

We work in a Martin-Löf type theory which has the following features.
\begin{itemize}
  \item Two universes named $\Set0$ and $\Set1$, where $\Set0$ supports
    inductive-recursive types (IR) as specified by Dybjer and Setzer
    \cite{dybjer99finite}. The $\Set1$ universe is not essential and we only use
    it as a convenience feature, in this paper and in the Agda formalization. We
    may omit the universe indices if they can be inferred or if we work over
    arbitrary indices.
  \item Function extensionality and uniqueness of identity proofs
    (UIP). Additionally, we assume equality reflection in this paper, thus
    working in extensional type theory, to avoid noise from equality transports.
  \item We write function types as $(x : A)\to B$ with $\lambda\,x.\,t$
    inhabitants, and $\Sigma$-types as $(x : A) \times B$, with pairing as
    $(t,\, u)$. We have $\top$ as the unit type with inhabitant $\tt$, and
    $\bot$ as the empty type. Propositional identity is written as $t = u$
    (coinciding with definitional equality).
  \item We occasionally use $\{x : A\} \to B$ for an Agda-like notation for
    function types with implicit arguments. We usually omit implicit
    applications but may explicitly write them as $t\,\{u\}$.  We may omit
    implicit function types altogether if it is clear where certain variables
    are quantified.
\end{itemize}

@@

\begin{mydefinition}
A \emph{category with family} (cwf) \cite{Dybjer96internaltype} consists of the following data:
\begin{itemize}
\item A category with a terminal object. We denote the set of objects as $\Con :
  \Seti$ and use capital Greek letters starting from $\Gamma$ to refer to
  objects. The set of morphisms is $\Sub : \Con \to \Con \to \Seti$, and we use
  $\sigma$, $\delta$ and so on to refer to morphisms. The terminal object is
  $\emptycon$ with unique morphism $\epsilon : \Sub\,\Gamma\,\emptycon$. In
  initial models (that is, syntaxes) of type theories, objects correspond to
  typing contexts, morphisms to parallel substitutions and the terminal object to
  the empty context; this informs the naming scheme that we use here.
\item A \emph{family structure}, containing $\Ty : \Con \to \Seti$ and $\Tm :
  (\Gamma : \Con) \to \Ty\,\Gamma \to \Seti$, where $\Ty$ is a presheaf over the
  category of contexts and $\Tm$ is a presheaf over the category of elements of
  $\Ty$. This means that both types ($\Ty$) and terms ($\Tm$) can be
  substituted, and substitution has functorial action. We use $A$, $B$, $C$ to
  refer to types and $t$, $u$, $v$ to refer to terms, and use $A[\sigma]$ and
  $t[\sigma]$ for substituting types and terms. Additionally, a family structure
  has \emph{context comprehension} which consists of a context extension
  operation $\blank\ext\blank : (\Gamma : \Con) \to \Ty\,\Gamma \to \Con$
  together with an isomorphism $\Sub\,\Gamma\,(\Delta\ext A) \simeq ((\sigma :
  \Sub\,\Gamma\,\Delta) \times \Tm\,\Gamma\,(A[\sigma]))$ which is natural in
  $\Gamma$.
\end{itemize}
\end{mydefinition}

@@

While the staging aspect could be subject of future research, we currently focus
on ``sizing'' hierarchies instead of staging hierarchies. This means that we
want to eliminate from any family to any other family which is ``connected'' by
a morphism.

@@

Unlike every other type or term former, there is no lifting computation rule for
$\El$ and $\Code$. Intuitively, the issue is that we would need to relate type
lifting and term lifting, but while term lifting is invertible, type lifting is
not. $\Lift$ sends a $\Ty_i\,\Gamma$ to a $\Ty_j\,\Gamma$, and a larger $\Ty$ is
not isomorphic a smaller one, because it contains more universes. So, for
example, lifting $\Bool_0 : \Ty_0\,\Gamma$ as a type to $\Ty_1\,\Gamma$ yields
$\Bool_1$, but lifting $\Bool_0$ as a term yields $\Bool_0$.

@@

\textbf{Bounded universe polymorphism} is realized by quantifying
over levels and morphisms with the usual $\Pi$ types. For example, if internal
levels strictly correspond to natural numbers, we may have
\begin{alignat*}{3}
  &\msf{idUpTo3} : \Pi(l : \Nat_3)(p : l <^I 3)(A : \U\,l\,3\,(\mkMor\,p)) \to \Lift\,(\mkMor\,p)\,A \to \Lift\,(\mkMor\,p)\,A\\
  &\msf{idUpTo3} = \lambda\,l\,p\,A\,a.\,a
\end{alignat*}
Here, we make sure that all types in the $\Pi$ are on the same level. We assume
that internal levels are in $\Nat_0$, but we can bind an $l : \Nat_3$, because
by cumulativity $l$ is also a term of $\Nat_0$.

@@

Note that $\Pi$ is contravariant in the domain. This is easily supported
with our IR semantics, unlike in the set-theoretic model of cumulativity for Coq
\cite{timany2018cumulative}, where function domains are invariant.

@@

Several related are not discussed in this paper and could be subject to future
work.

First, besides consistency, we are often interested in \emph{canonicity},
\emph{normalization} or other metatheoretical properties. The current work
focuses on consistency and leaves other properties to future work. We did keep
canonicity in mind when specifying the systems in this paper. Hopefully the
usual proof method of gluing (in other words, proof-relevant logical predicates)
\cite{kaposi2019gluing, sterling2019algebraic} can be adapted to the theories in
this paper. It seems that the main task in doing so is to define proof-relevant
logical predicates over $\uir$.

Second, we only focus on using universes as sized-based classifiers for
types. Stratification features are also present in two-level type theory
\cite{twolevel}, modal type theories \cite{gratzer20multimodal} or as h-levels
in homotopy type theory \cite{hottbook}. It would be interesting to port
universe features in this paper to two-level type theory, as they would
hopefully model a form of stage polymorphism in multi-stage compilation. It
would be nice to try representing $\msf{Prop}$ universes in TTFL as well;
this is closed related to h-level based stratification.

@@

Lastly, we do not handle impredicative universes. The main reason for this is
that we do not know the consistency of having IR and impredicative function
space together in the same universe, and modeling impredicativity seems to
require this assumption in the metatheory. This could be investigated as well
in future research.

>>>>

Users of type theories often view universe levels as a bureaucratic detail, a
necessary annoyance in service of boosting expressive power while retaining
logical consistency. However, universe hierarchies are not going away any time
soon in practical implementations of type theory. In recent developments of
systems, we are getting more universes and more adjacent features:
\begin{itemize}
\item Agda recently added a limited cumulativity as an optional feature
  for universes \cite{agdadocs}, and the upcoming 2.6.2 version will extend the $\omega+1$
  universe hierarchy to $\omega*2$.
\item Coq added support for cumulative inductive types \cite{timany18cumulative}
  and a form of bounded universe polymorphism \cite{ziliani15unification}.
\end{itemize}
\noindent At this point, there is a veritable zoo of universe features in existing
implementations. We have perhaps even more design choices when considering the
formal metatheory of type theories. Do type formers stay in the same
universe, or take the $\sqcup$ of universes of constituent types? Can
eliminators target any universe, or do we instead use lifting operators to cross
levels? What kind of universe polymorphism do we have, can we quantify over
level bounds? Is there a type of levels, or are levels in a separate syntactic layer?

@@

We work in a Martin-Löf type theory which has the following features.
\begin{itemize}
  \item Two universes named $\Set0$ and $\Set1$, where $\Set0$ supports
    inductive-recursive types (IR) as specified by Dybjer and Setzer
    \cite{dybjer99finite}. The $\Set1$ universe is not essential and we only use
    it as a convenience feature, in this paper and in the Agda formalization. We
    may omit the universe indices if they can be inferred or if we work over
    arbitrary indices.
  \item Function extensionality and uniqueness of identity proofs
    (UIP). Additionally, we assume equality reflection in this paper, thus
    working in extensional type theory, to avoid noise from equality transports.
  \item We write function types as $(x : A)\to B$ with $\lambda\,x.\,t$
    inhabitants, and $\Sigma$-types as $(x : A) \times B$, with pairing as
    $(t,\, u)$. We have $\top$ as the unit type with inhabitant $\tt$, $\bot$ as
    the empty type, and $\Bool$ with $\true$ and $\false$
    inhabitants. Propositional identity is written as $t = u$ (coinciding with
    definitional equality).
  \item We occasionally use $\{x : A\} \to B$ for an Agda-like notation for
    function types with implicit arguments. We usually omit implicit
    applications but may explicitly write them as $t\,\{u\}$.  We may omit
    implicit function types altogether if it is clear where certain variables
    are quantified.
\end{itemize}

@@

\begin{mydefinition}
A \emph{category with family} (cwf) \cite{Dybjer96internaltype} consists of the following data:
\begin{itemize}
\item A category with a terminal object. We denote the set of objects as $\Con :
  \Seti$ and use capital Greek letters starting from $\Gamma$ to refer to
  objects. The set of morphisms is $\Sub : \Con \to \Con \to \Seti$, and we use
  $\sigma$, $\delta$ and so on to refer to morphisms. The terminal object is
  $\emptycon$ with unique morphism $\epsilon : \Sub\,\Gamma\,\emptycon$. In
  initial models (that is, syntaxes) of type theories, objects correspond to
  typing contexts, morphisms to parallel substitutions and the terminal object to
  the empty context; this informs the naming scheme.
\item A \emph{family structure}, containing $\Ty : \Con \to \Seti$ and $\Tm :
  (\Gamma : \Con) \to \Ty\,\Gamma \to \Seti$, where $\Ty$ is a presheaf over the
  category of contexts and $\Tm$ is a presheaf over the category of elements of
  $\Ty$. This means that both types ($\Ty$) and terms ($\Tm$) can be
  substituted, and substitution has functorial action. We use $A$, $B$, $C$ to
  refer to types and $t$, $u$, $v$ to refer to terms, and use $A[\sigma]$ and
  $t[\sigma]$ for substituting types and terms. Additionally, a family structure
  has \emph{context comprehension} which consists of a context extension
  operation $\blank\ext\blank : (\Gamma : \Con) \to \Ty\,\Gamma \to \Con$
  together with an isomorphism $\Sub\,\Gamma\,(\Delta\ext A) \simeq ((\sigma :
  \Sub\,\Gamma\,\Delta) \times \Tm\,\Gamma\,(A[\sigma]))$ which is natural in
  $\Gamma$.
\end{itemize}
\end{mydefinition}

@@

While the staging aspect could be subject of future research, we currently focus
on ``sizing'' hierarchies instead of staging hierarchies. This means that we
want to eliminate from any family to any other family which is connected by
a morphism.

@@

Unlike every other type or term former, there is no lifting computation rule for
$\El$ and $\Code$. Intuitively, the issue is that we would need to relate type
lifting and term lifting, but while term lifting is invertible, type lifting is
not. $\Lift$ sends a $\Ty_i\,\Gamma$ to a $\Ty_j\,\Gamma$, and $\Ty_j\,\Gamma$
is not isomorphic to $\Ty_i\,\Gamma$, because it contains more universes. So,
for example, lifting $\Bool_0 : \Ty_0\,\Gamma$ as a type to $\Ty_1\,\Gamma$
yields $\Bool_1$, but lifting $\Bool_0$ as a term yields $\Bool_0$.

@@

\textbf{Bounded universe polymorphism} is realized by quantifying over levels
and morphisms with the usual $\Pi$ types. For example, if levels strictly
correspond to internal natural numbers, we may have
\begin{alignat*}{3}
  &\msf{idUpTo3} : \Pi(l : \Nat_3)(p : \Lift_0^3(l <^I 3))(A : \U\,l\,3\,(\mkMor\,p)) \to \Lift\,(\mkMor\,p)\,A \to \Lift\,(\mkMor\,p)\,A\\
  &\msf{idUpTo3} = \lambda\,l\,p\,A\,a.\,a
\end{alignat*}
Here, we make sure that all types are on the same level, by appropriate
lifting. We assume that internal levels are in $\Nat_0$, but we can bind an $l :
\Nat_3$, because by cumulativity $l$ is also a term of $\Nat_0$. Likewise, the
$p$ variable is a term of $\Lift_0^3(l <^I 3)$ and $l <^I 3$ as well.

@@

Note that $\Pi$ is contravariant in the domain. This is easily supported
with our IR semantics, unlike in the set-theoretic model of cumulativity for Coq
\cite{timany18cumulative}, where function domains are invariant.

@@

Several related topics are not discussed in this paper and could be subject to
future work.

First, besides consistency, we are often interested in \emph{canonicity},
\emph{normalization} or other metatheoretical properties. The current work
focuses on consistency and leaves other properties to future work. We did keep
canonicity in mind when specifying the systems in this paper. Hopefully the
usual proof method of gluing (in other words, proof-relevant logical predicates)
\cite{kaposi2019gluing, sterling2019algebraic} can be adapted to the theories in
this paper.

Second, we only focus on using universes as sized-based classifiers for
types. Stratification features are also present in two-level type theory
\cite{twolevel}, modal type theories \cite{gratzer20multimodal} or as h-levels
in homotopy type theory \cite{hottbook}. It would be interesting to port
universe features in this paper to two-level type theory, as they would
hopefully model a form of stage polymorphism in multi-stage compilation. We
could try representing $\msf{Prop}$ universes in TTFL as well. This is closely
related to h-level based stratification.

@@

Lastly, we do not handle impredicative universes. The main reason for this is
that we do not know the consistency of having IR and impredicative function
space together in the same universe, and modeling impredicativity seems to
require this assumption in the metatheory. This could be investigated as well
in future work.
